---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
set.seed(42)
```
# dqsample

The 'base::sample' function uses a slightly biased algorithm for
creating random integers within a given range. As an alternative the
algorithm suggested by Daniel Lemire (2018,
<[arXiv:1805.1094](https://arxiv.org/abs/1805.10941)>) is used.

## Installation

At the moment dqsample is not on CRAN, but you can install the current version
via [drat](https://cran.r-project.org/package=drat):

```r
if (!requireNamespace("drat", quietly = TRUE)) install.packages("drat")
drat::addRepo("daqana")
install.packages("dqsample")
```

## Example

When sampling many random integers the number of odd and even numbers should be about the same.
This is not the case with `base::sample`:

```{r base}
m <- 2/5 * 2^32
x <- base::sample(m, 1000000, replace = TRUE)
table(x %% 2)
```

While it is the case with `dqsample::sample`:

```{r dqsample}
m <- 2/5 * 2^32
x <- dqsample::sample(m, 1000000, replace = TRUE)
table(x %% 2)
```
