---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

[![Travis build status](https://travis-ci.org/daqana/dqsample.svg?branch=master)](https://travis-ci.org/daqana/dqsample)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/daqana/dqsample?branch=master&svg=true)](https://ci.appveyor.com/project/rstub/dqsample)
[![CRAN status](https://www.r-pkg.org/badges/version/dqsample)](https://cran.r-project.org/package=dqsample)
[![Coverage status](https://codecov.io/gh/daqana/dqsample/branch/master/graph/badge.svg)](https://codecov.io/github/daqana/dqsample?branch=master)

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
set.seed(42)
```
# dqsample

The `base::sample` function uses a slightly biased algorithm for
creating random integers within a given range. As an alternative the
algorithm suggested by Daniel Lemire (2018,
<[arXiv:1805.1094](https://arxiv.org/abs/1805.10941)>) is used. 
The package is motivated by a [thread on R-devel](https://stat.ethz.ch/pipermail/r-devel/2018-September/076817.html).

## Installation

At the moment dqsample is not on CRAN, but you can install the current version
via [drat](https://cran.r-project.org/package=drat):

```r
if (!requireNamespace("drat", quietly = TRUE)) install.packages("drat")
drat::addRepo("daqana")
install.packages("dqsample")
```

## Example

When sampling many random integers the density of odd and even numbers should be roughly equal and constant.
This is not the case with `base::sample`:

```{r base}
m <- 2/5 * 2^32
x <- base::sample(floor(m), 1000000, replace = TRUE)
plot(density(x[x %% 2 == 0]), main = "base::sample")
lines(density(x[x %% 2 == 1]), col = 2)
```

While it is the case with `dqsample::sample`:

```{r dqsample}
m <- 2/5 * 2^32
x <- dqsample::sample(floor(m), 1000000, replace = TRUE)
plot(density(x[x %% 2 == 0]), main = "dqsample::sample")
lines(density(x[x %% 2 == 1]), col = 2)
```

This particular sample for the bias was found by [Duncan Murdoch](https://stat.ethz.ch/pipermail/r-devel/2018-September/076827.html).
